IGAS Project

### 2023-10-09 IGAS.qmd
update citations

---
title: IGAS Project
date: last-modified
#page-layout: full
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: false 
---

@@

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer to novel tasks or conditions  (Barnett & Ceci, 2002; Detterman, 1993). Such transfer challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence transfer, and to develop cognitive models that can predict when transfer is likely to occur. The factor of interest to the present investigation is variation during training. Our experiments add to the longstanding empirical investigation of the controversial relationship between training variation, and subsequent transfer. We also offer a novel explanation for such results in the form of an instance-based model that accounts for the benefits of variation in simple terms of psychological similarity. We first review the relevant concepts and literature.

@@

Notions of similarity have long played a central role in many prominent models of generalization of learning, as well as in the longstanding theoretical issue of whether learners abstract an aggregate, summary representation, or if they simply store individual instances. Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar (Posner & Keele, 1968).  Prototype models were later challenged by the success of instance-based or exemplar models – which were shown to provide an account of generalization as good or better than prototype models, with the advantage of not assuming the explicit construction of an internal prototype (Estes, 1994; Hintzman, 1984; Medin & Schaffer, 1978; Nosofsky, 1986). Instance-based models assume that learners encode each experience with a task as a separate instance/exemplar/trace, and that each encoded trace is in turn compared against novel stimuli. As the number of stored instances increases, so does the likelihood that some previously stored instance will be retrieved to aid in the performance of a novel task. Stored instances are retrieved in the context of novel stimuli or tasks if they are sufficiently similar, thus suggesting that the process of computing similarity is of central importance to generalization.

Similarity, defined in this literature as a function of psychological distance between instances or categories, has provided a successful account of generalization across numerous tasks and domains. In an influential study demonstrating an ordinal similarity effect, experimenters employed a numerosity judgment task in which participants quickly report the number of dots flashed on a screen. Performance (in terms of response times to new patterns) on novel dot configurations varied as an inverse function of their similarity to previously trained dot configurations (Palmeri, 1997). That is, performance was better on novel configurations moderately similar to trained configurations than to configurations with low-similarity, and also better on low-similarity configurations than to even less similar, unrelated configurations. Instance-based approaches have had some success accounting for performance in certain sub-domains of motor learning (Cohen & Rosenbaum, 2004; Crump & Logan, 2010; Meigh et al., 2018; Poldrack et al., 1999; Wifall et al., 2017) . Crump and Logan (2010) trained participants to type words on an unfamiliar keyboard, while constraining the letters composing the training words to a pre-specified letter set. Following training, typing speed was tested on previously experienced words composed of previously experienced letters; novel words composed of letters from the trained letter set; and novel words composed of letters from an untrained letter set. Consistent with an instance-based account, transfer performance was graded such that participants were fastest at typing the words they had previously trained on, followed by novel words composed of letters they had trained on, and slowest performance for new words composed of untrained letters.

@@

While similarity-based models account for transfer by the degree of similarity between previous and new experiences, a largely separate body of research has focused on improving transfer by manipulating characteristics of the initial training stage. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work – variability of training examples.

Research on the effects of varied training typically compares participants trained under constant, or minimal variability conditions to those trained from a variety of examples or conditions (Czyż, 2021; Soderstrom & Bjork, 2015). Varied training has been shown to influence learning in myriad domains including categorization of simple stimuli (Hahn et al., 2005; Maddox & Filoteo, 2011; Posner & Keele, 1968), complex categorization (Nosofsky et al., 2018), language learning (Jones & Brandt, 2020; Perry et al., 2010; Twomey et al., 2018; Wonnacott et al., 2012) anagram completion (Goode et al., 2008), trajectory extrapolation (Fulvio et al., 2014), task switching (Sabah et al., 2019), associative learning (Lee et al., 2019), visual search (George & Egner, 2021; Gonzalez & Madhavan, 2011; Kelley & Yantis, 2009), voice identity learning (Lavan et al., 2019), simple motor learning (Braun et al., 2009; Kerr & Booth, 1978; Roller et al., 2001; Willey & Liu, 2018), sports training (Green et al., 1995; North et al., 2019), and training on a complex video game (Seow et al., 2019).

@@

### Sample Size Estimation
To obtain an independent estimate of effect size, we identified previous investigations which included between-subjects contrasts of varied and constant conditions following training on an accuracy based projectile launching task (Chua et al., 2019; Goodwin et al., 1998; Kerr & Booth, 1978; Wulf, 1991). We then averaged effects across these studies, yielding a Cohens f =.43. The GPower 3.1 software package (Faul et al., 2009) was then used to determine that a power of 80% requires a sample size of at least 23 participants per condition. All experiments reported in the present manuscript exceed this minimum number of participants per condition.

### Participants 
Participants were recruited from an undergraduate population that is 63% female and consists almost entirely of individuals aged 18-22 years. A total of 110 Indiana University psychology students participated in Experiment 1. We subsequently excluded 34 participants poor performance at one of the dependent measures of the task (2.5-3 standard deviations worse than the median subject at the task) or for displaying a pattern of responses that was clearly indicative of a lack of engagement with the task (e.g. simply dropping the ball on each trial rather than throwing it at the target), or for reporting that they completed the experiment on a phone or tablet device, despite the instructions not to use one of these devices. A total of 74 participants were retained for the final analyses, 35 in the varied group and 39 in the constant group.

### Task
The experimental task was programmed in JavaScript, using packages from the Phaser physics engine (https://phaser.io) and the jsPsych library (de Leeuw, 2015). The stimuli, presented on a black background, consisted of a circular blue ball – controlled by the participant via the mouse or trackpad cursor; a rectangular green target; a red rectangular barrier located between the ball and the target; and an orange square within which the participant could control the ball before releasing it in a throw towards the target. Because the task was administered online, the absolute distance between stimuli could vary depending on the size of the computer monitor being used, but the relative distance between the stimuli was held constant. Likewise, the distance between the center of the target, and the training and testing locations was scaled such that relative distances were preserved regardless of screen size. For the sake of brevity, subsequent mentions of this relative distance between stimuli, or the position where the ball landed in relation to the center of the target, will be referred to simply as distance. Methods Figure 2 displays the layout of the task, as it would appear to a participant at the start of a trial, with the ball appearing in the center of the orange square. 
Using a mouse or trackpad, participants click down on the ball to take control of the ball, connecting the movement of the ball to the movement of the cursor. Participants can then “wind up” the ball by dragging it (within the confines of the orange square) and then launch the ball by releasing the cursor. If the ball does not land on the target, participants are presented with feedback in red text at the top right of the screen, on how many units away they were from the center of the target. If the ball was thrown outside of the boundary of the screen participants are given feedback as to how far away from the target center the ball would have been if it had continued its trajectory. If the ball strikes the barrier (from the side or by landing on top), feedback is presented telling participants to avoid hitting the barrier. If participants drag the ball outside of the orange square before releasing it, the trial terminates, and they are reminded to release the ball within the orange square. If the ball lands on the target, feedback is presented in green text, confirming that the target was hit, and presenting additional feedback on how many units away the ball was from the exact center of the target.

@@

To prepare the data, we first removed trials that were not easily interpretable as performance indicators in our task. Removed trials included: 1) those in which participants dragged the ball outside of the orange starting box without releasing it, 2) trials in which participants clicked on the ball, and then immediately released it, causing the ball to drop straight down, 3) outlier trials in which the ball was thrown more than 2.5 standard deviations further than the average throw (calculated separately for each throwing position), and 4) trials in which the ball struck the barrier. 
The primary measure of performance used in all analyses was the absolute distance away from the center of the target. The absolute distance was calculated on every trial, and then averaged within each subject to yield a single performance score, for each position. A consistent pattern across training and testing phases in both experiments was for participants to perform worse from throwing positions further away from the target – a pattern which we refer to as the difficulty of the positions. However, there were no interactions between throwing position and training conditions, allowing us to collapse across positions in cases where contrasts for specific positions were not of interest. All data processing and statistical analyses were performed in R version 4.03 (R Core Team, 2020). ANOVAs for group comparisons were performed using the rstatix package (Kassambara, 2021).

@@

We performed an ANOVA comparison with stage as a within-group factor and condition as between-group factor. The analysis revealed a significant effect of training stage F(2,142)=62.4, p<.001, $\eta^{2}_G$ = .17, such that performance improved over the course of training There was no significant effect of condition F(1,71)=1.42, p=.24, $\eta^{2}_G$ = .02,  and no significant interaction between condition and training stage, F(2,142)=.10, p=.91, $\eta^{2}_G$ < .01.

@@

## Testing Phase 
In Experiment 1, a single constant-trained group was compared against a single varied-trained group. At the transfer phase, all participants were tested from 3 positions: 1) the positions(s) from their own training, 2) the training position(s) of the other group, and 3) a position novel to both groups. Overall, group performance was compared with a mixed type III ANOVA, with condition (varied vs. constant) as a between-subject factor and throwing location as a within-subject variable. The effect of throwing position was strong, F(3,213) = 56.12, p<.001, η2G = .23. The effect of training condition was significant F(1,71)=8.19, p<.01, η2G = .07. There was no significant interaction between group and position, F(3,213)=1.81, p=.15, η2G = .01.

@@

### Participants 
A total of 306 Indiana University psychology students participated in Experiment 2, which was also conducted online. As was the case in experiment 1, the undergraduate population from which we recruited participants was 63% female and primarily composed of 18–22-year-old individuals. Using the same procedure as experiment 1, we excluded 98 participants for exceptionally poor performance at one of the dependent measures of the task, or for displaying a pattern of responses indicative of a lack of engagement with the task. A total of 208 participants were included in the final analyses with 31 in the varied group and 32, 28, 37, 25, 29, 26 participants in the constant groups training from location 400, 500, 625, 675, 800, and 900, respectively. All participants were compensated with course credit.

### Task and Procedure 
The task of Experiment 2 was identical to that of Experiment 1, in all but some minor adjustments to the height of the barrier, and the relative distance between the barrier and the target. Additionally, the intermittent testing trials featured in experiment 1 were not utilized in experiment 2, and all training and testing trials were presented with feedback. An abbreviated demo of the task used for Experiment 2 can be found at (https://pcl.sitehost.iu.edu/tg/demos/igas_expt2_demo.html).

@@

### Data Processing and Statistical Packages  
After confirming that condition and throwing position did not have any significant interactions, we standardized performance within each position, and then average across position to yield a single performance measure per participant. This standardization did not influence our pattern of results. As in experiment 1, we performed type III ANOVA’s due to our unbalanced design, however the pattern of results presented below is not altered if type 1 or type III tests are used instead.  The statistical software for the primary analyses was the same as for experiment 1. Individual learning rates in the testing phase, compared between groups in the supplementary analyses, were fit using the TEfit package in R (Cochrane, 2020).

### Training Phase 
The different training conditions trained from positions that were not equivalently difficult and are thus not easily amenable to comparison. As previously stated, the primary interest of the training data is confirmation that some learning did occur. Figure 5 depicts the training performance of the varied group alongside that of the aggregate of the six constant groups (5a), and each of the 6 separate constant groups (5b). An ANOVA comparison with training stage (beginning, middle, end) as a within-group factor and group (the varied condition vs. the 6 constant conditions collapsed together) as a between-subject factor revealed no significant effect of group on training performance, F(1,206)=.55,p=.49, $\eta^{2}_G$ <.01, a significant effect of training stage F(2,412)=77.91, p<.001, $\eta^{2}_G$ =.05, and no significant interaction between group and training stage, F(2,412)=.489 p=.61, $\eta^{2}_G$ <.01. We also tested for a difference in training performance between the varied group and the two constant groups that trained matching throwing positions (i.e., the constant groups training from position 500, and position 800). The results of our ANOVA on this limited dataset mirrors that of the full-group analysis, with no significant effect of group F(1,86)=.48, p=.49, $\eta^{2}_G$ <.01, a significant effect of training stage F(2,172)=56.29, p<.001, $\eta^{2}_G$ =.11, and no significant interaction between group and training stage, F(2,172)=.341 p=.71, $\eta^{2}_G$ <.01.

@@

### Testing Phase
In Experiment 2, a single varied condition (trained from two positions, 500 and 800), was compared against six separate constant groups (trained from a single position, 400, 500, 625, 675, 800 or 900). For the testing phase, all participants were tested from all six positions, four of which were novel for the varied condition, and five of which were novel for each of the constant groups. For a general comparison, we took the absolute deviations for each throwing position and computed standardized scores across all participants, and then averaged across throwing position. The six constant groups were then collapsed together allowing us to make a simple comparison between training conditions (constant vs. varied). A type III between-subjects ANOVA was performed, yielding a significant effect of condition F(1,206)=4.33, p=.039, $\eta^{2}_G$ =.02. Descriptive statistics for each condition are shown in table 2. Figure 6A visualizes the consistent advantage of the varied condition over the constant groups across the testing positions. Figure 6b shows performance between the varied condition and the individual constant groups.

@@

## Discussion 
The results of experiment 2 largely conform to the findings of experiment 1. Participants in both varied and constant conditions improved at the task during the training phase. We did not observe the common finding of training under varied conditions producing worse performance during acquisition than training under constant conditions (Catalano & Kleiner, 1984; Wrisberg et al., 1987), which has been suggested to relate to the subsequent benefits of varied training in retention and generalization testing (Soderstrom & Bjork, 2015). However our finding of no difference in training performance between constant and varied groups has been observed in previous work (Chua et al., 2019; Moxley, 1979; Pigott & Shapiro, 1984).

In the testing phase, our varied group significantly outperformed the constant conditions in both a general comparison, and in an analysis limited to novel throwing positions. The observed benefit of varied over constant training echoes the findings of many previous visuomotor skill learning studies that have continued to emerge since the introduction of Schmidt’s influential Schema Theory (Catalano & Kleiner, 1984; Chua et al., 2019; Goodwin et al., 1998; McCracken & Stelmach, 1977; Moxley, 1979; Newell & Shapiro, 1976; Pigott & Shapiro, 1984; Roller et al., 2001; Schmidt, 1975; Willey & Liu, 2018; Wrisberg et al., 1987; Wulf, 1991). We also join a much smaller set of research to observe this pattern in a computerized task (Seow et al., 2019). One departure from the experiment 1 findings concerns the pattern wherein the varied group outperformed the constant group even from the training position of the constant group, which was significant in experiment 1, but did not reach significance in experiment 2. Although this pattern has been observed elsewhere in the literature (Goode et al., 2008; Kerr & Booth, 1978), the overall evidence for this effect appears to be far weaker than for the more general benefit of varied training in conditions novel to all training groups.

# Computational Model
Controlling for the similarity between training and testing
The primary goal of Experiment 2 was to examine whether the benefits of variability would persist after accounting for individual differences in the similarity between trained and tested throwing locations. To this end, we modelled each throw as a two-dimensional point in the space of x and y velocities applied to the projectile at the moment of release. For each participant, we took each individual training throw, and computed the similarity between that throw and the entire population of throws within the solution space for each of the 6 testing positions. We defined the solution space empirically as the set of all combinations of x and y throw velocities that resulted in hitting the target. We then summed each of the trial-level similarities to produce a single similarity for each testing position score relating how the participant threw the ball during training and the solutions that would result in target hits from each of the six testing positions – thus resulting in six separate similarity scores for each participant. Figure 8a visualizes the solution space for each location and illustrates how different combinations of x and y velocity result in successfully striking the target from different launching positions. As illustrated in Figure 8b, the solution throws represent just a small fraction of the entire space of velocity combinations used by participants throughout the experiment.

@@

For each individual trial, the Euclidean distance (Equation 1) was computed between the velocity components (x and y) of that trial and the velocity components of each individual solution throw for each of the 6 positions from which participants would be tested in the final phase of the study.  The P parameter in Equation 1 is set equal to 2, reflecting a Gaussian similarity gradient. Then, as per an instance-based model of similarity (Logan, 2002; Nosofsky, 1992), these distances were multiplied by a sensitivity parameter, c, and then exponentiated to yield a similarity value. The parameter c controls the rate with which similarity-based generalization drops off as the Euclidean distance between two throws in x- and y-velocity space increases.  If c has a large value, then even a small difference between two throws’ velocities greatly decreases the extent of generalization from one to the other. A small value for c produces broad generalization from one throw to another despite relatively large differences in their velocities. The similarity values for each training individual throw made by a given participant were then summed to yield a final similarity score, with a separate score computed for each of the 6 testing positions. The final similarity score is construable as index of how accurate the throws a participant made during the training phase would be for each of the testing positions.

**Equation 1:**
\[ Similarity_{I,J} =  \sum_{i = I}\sum_{j=J} e^{-c\cdot d^p_{i,j}} \]

@@

## Conclusion
Our experiments demonstrate a reliable benefit of varied training in a simple projectile launching task. Such results were accounted for by an instance-based model that assumes that varied training results in the computation of a broader similarity-based generalization gradient. Instance-based models augmented with this assumption may be a valuable approach towards better understanding skill generalization and transfer.

>>>>

---
title: IGAS Project
date: last-modified
#page-layout: full
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: false 
editor: 
  markdown: 
    wrap: 72
---

@@

The past century of research on human learning has produced ample
evidence that although learners can improve at almost any task, such
improvements are often specific to the trained task, with unreliable or
even nonexistent transfer to novel tasks or conditions
[@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such
transfer challenges are of noteworthy practical relevance, given that
educators, trainers, and rehabilitators typically intend for their
students to be able to apply what they have learned to new situations.
It is therefore important to better understand the factors that
influence transfer, and to develop cognitive models that can predict
when transfer is likely to occur. The factor of interest to the present
investigation is variation during training. Our experiments add to the
longstanding empirical investigation of the controversial relationship
between training variation, and subsequent transfer. We also offer a
novel explanation for such results in the form of an instance-based
model that accounts for the benefits of variation in simple terms of
psychological similarity. We first review the relevant concepts and
literature.

@@

Notions of similarity have long played a central role in many prominent
models of generalization of learning, as well as in the longstanding
theoretical issue of whether learners abstract an aggregate, summary
representation, or if they simply store individual instances. Early
models of learning often assumed that discrete experiences with some
task or category were not stored individually in memory, but instead
promoted the formation of a summary representation, often referred to as
a prototype or schema, and that exposure to novel examples would then
prompt the retrieval of whichever preexisting prototype was most similar
@posnerGenesisAbstractIdeas1968 [Posner & Keele, 1968]. Prototype
models were later challenged by the success of instance-based or
exemplar models -- which were shown to provide an account of
generalization as good or better than prototype models, with the
advantage of not assuming the explicit construction of an internal
prototype @estesArrayModelsCategory1986;
@hintzmanMINERVASimulationModel1984;
@medinContextTheoryClassification1978;
@nosofskyAttentionSimilarityIdentificationcategorization1986 ].
Instance-based models assume that learners encode each experience with a
task as a separate instance/exemplar/trace, and that each encoded trace
is in turn compared against novel stimuli. As the number of stored
instances increases, so does the likelihood that some previously stored
instance will be retrieved to aid in the performance of a novel task.
Stored instances are retrieved in the context of novel stimuli or tasks
if they are sufficiently similar, thus suggesting that the process of
computing similarity is of central importance to generalization.

Similarity, defined in this literature as a function of psychological
distance between instances or categories, has provided a successful
account of generalization across numerous tasks and domains. In an
influential study demonstrating an ordinal similarity effect,
experimenters employed a numerosity judgment task in which participants
quickly report the number of dots flashed on a screen. Performance (in
terms of response times to new patterns) on novel dot configurations
varied as an inverse function of their similarity to previously trained
dot configurations @palmeriExemplarSimilarityDevelopment1997. That is, performance was better on
novel configurations moderately similar to trained configurations than
to configurations with low-similarity, and also better on low-similarity
configurations than to even less similar, unrelated configurations.
Instance-based approaches have had some success accounting for
performance in certain sub-domains of motor learning [@cohenWhereGraspsAre2004; @crumpEpisodicContributionsSequential2010; @meighWhatMemoryRepresentation2018; @poldrackRelationshipSkillLearning1999]; @wifallPerceptualSimilarityAffects.@crumpEpisodicContributionsSequential2010 trained participants to type words on an unfamiliar keyboard, while constraining the letters
composing the training words to a pre-specified letter set. Following
training, typing speed was tested on previously experienced words
composed of previously experienced letters; novel words composed of
letters from the trained letter set; and novel words composed of letters
from an untrained letter set. Consistent with an instance-based account,
transfer performance was graded such that participants were fastest at
typing the words they had previously trained on, followed by novel words
composed of letters they had trained on, and slowest performance for new
words composed of untrained letters.

@@

While similarity-based models account for transfer by the degree of
similarity between previous and new experiences, a largely separate body
of research has focused on improving transfer by manipulating
characteristics of the initial training stage. Such characteristics have
included training difficulty, spacing, temporal order, feedback
schedules, and the primary focus of the current work -- variability of
training examples.

Research on the effects of varied training typically compares
participants trained under constant, or minimal variability conditions
to those trained from a variety of examples or conditions [@czyzVariabilityPracticeInformation2021; @soderstromLearningPerformanceIntegrative2015]. Varied training has been shown to influence
learning in myriad domains including categorization of simple stimuli [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968],
complex categorization [@nosofskyModelguidedSearchOptimal2018], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory
extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice
identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018],
sports training  [@greenPracticeVariabilityTransfer1995; @northEffectConsistentVaried2019], and training
on a complex video game [@seowTransferEffectsVaried2019].

@@

### Sample Size Estimation

To obtain an independent estimate of effect size, we identified previous
investigations which included between-subjects contrasts of varied and
constant conditions following training on an accuracy based projectile
launching task (Chua et al., 2019; Goodwin et al., 1998; Kerr & Booth,
1978; Wulf, 1991). We then averaged effects across these studies,
yielding a Cohens f =.43. The GPower 3.1 software package (Faul et al.,
2009) was then used to determine that a power of 80% requires a sample
size of at least 23 participants per condition. All experiments reported
in the present manuscript exceed this minimum number of participants per
condition.

### Participants

Participants were recruited from an undergraduate population that is 63%
female and consists almost entirely of individuals aged 18-22 years. A
total of 110 Indiana University psychology students participated in
Experiment 1. We subsequently excluded 34 participants poor performance
at one of the dependent measures of the task (2.5-3 standard deviations
worse than the median subject at the task) or for displaying a pattern
of responses that was clearly indicative of a lack of engagement with
the task (e.g. simply dropping the ball on each trial rather than
throwing it at the target), or for reporting that they completed the
experiment on a phone or tablet device, despite the instructions not to
use one of these devices. A total of 74 participants were retained for
the final analyses, 35 in the varied group and 39 in the constant group.

### Task

The experimental task was programmed in JavaScript, using packages from
the Phaser physics engine (https://phaser.io) and the jsPsych library
(de Leeuw, 2015). The stimuli, presented on a black background,
consisted of a circular blue ball -- controlled by the participant via
the mouse or trackpad cursor; a rectangular green target; a red
rectangular barrier located between the ball and the target; and an
orange square within which the participant could control the ball before
releasing it in a throw towards the target. Because the task was
administered online, the absolute distance between stimuli could vary
depending on the size of the computer monitor being used, but the
relative distance between the stimuli was held constant. Likewise, the
distance between the center of the target, and the training and testing
locations was scaled such that relative distances were preserved
regardless of screen size. For the sake of brevity, subsequent mentions
of this relative distance between stimuli, or the position where the
ball landed in relation to the center of the target, will be referred to
simply as distance. Methods Figure 2 displays the layout of the task, as
it would appear to a participant at the start of a trial, with the ball
appearing in the center of the orange square. Using a mouse or trackpad,
participants click down on the ball to take control of the ball,
connecting the movement of the ball to the movement of the cursor.
Participants can then "wind up" the ball by dragging it (within the
confines of the orange square) and then launch the ball by releasing the
cursor. If the ball does not land on the target, participants are
presented with feedback in red text at the top right of the screen, on
how many units away they were from the center of the target. If the ball
was thrown outside of the boundary of the screen participants are given
feedback as to how far away from the target center the ball would have
been if it had continued its trajectory. If the ball strikes the barrier
(from the side or by landing on top), feedback is presented telling
participants to avoid hitting the barrier. If participants drag the ball
outside of the orange square before releasing it, the trial terminates,
and they are reminded to release the ball within the orange square. If
the ball lands on the target, feedback is presented in green text,
confirming that the target was hit, and presenting additional feedback
on how many units away the ball was from the exact center of the target.

@@

To prepare the data, we first removed trials that were not easily
interpretable as performance indicators in our task. Removed trials
included: 1) those in which participants dragged the ball outside of the
orange starting box without releasing it, 2) trials in which
participants clicked on the ball, and then immediately released it,
causing the ball to drop straight down, 3) outlier trials in which the
ball was thrown more than 2.5 standard deviations further than the
average throw (calculated separately for each throwing position), and 4)
trials in which the ball struck the barrier. The primary measure of
performance used in all analyses was the absolute distance away from the
center of the target. The absolute distance was calculated on every
trial, and then averaged within each subject to yield a single
performance score, for each position. A consistent pattern across
training and testing phases in both experiments was for participants to
perform worse from throwing positions further away from the target -- a
pattern which we refer to as the difficulty of the positions. However,
there were no interactions between throwing position and training
conditions, allowing us to collapse across positions in cases where
contrasts for specific positions were not of interest. All data
processing and statistical analyses were performed in R version 4.03 (R
Core Team, 2020). ANOVAs for group comparisons were performed using the
rstatix package (Kassambara, 2021).

@@

We performed an ANOVA comparison with stage as a within-group factor and
condition as between-group factor. The analysis revealed a significant
effect of training stage F(2,142)=62.4, p\<.001, $\eta^{2}_G$ = .17,
such that performance improved over the course of training There was no
significant effect of condition F(1,71)=1.42, p=.24, $\eta^{2}_G$ = .02,
and no significant interaction between condition and training stage,
F(2,142)=.10, p=.91, $\eta^{2}_G$ \< .01.

@@

## Testing Phase

In Experiment 1, a single constant-trained group was compared against a
single varied-trained group. At the transfer phase, all participants
were tested from 3 positions: 1) the positions(s) from their own
training, 2) the training position(s) of the other group, and 3) a
position novel to both groups. Overall, group performance was compared
with a mixed type III ANOVA, with condition (varied vs. constant) as a
between-subject factor and throwing location as a within-subject
variable. The effect of throwing position was strong, F(3,213) = 56.12,
p\<.001, η2G = .23. The effect of training condition was significant
F(1,71)=8.19, p\<.01, η2G = .07. There was no significant interaction
between group and position, F(3,213)=1.81, p=.15, η2G = .01.

@@

### Participants

A total of 306 Indiana University psychology students participated in
Experiment 2, which was also conducted online. As was the case in
experiment 1, the undergraduate population from which we recruited
participants was 63% female and primarily composed of 18--22-year-old
individuals. Using the same procedure as experiment 1, we excluded 98
participants for exceptionally poor performance at one of the dependent
measures of the task, or for displaying a pattern of responses
indicative of a lack of engagement with the task. A total of 208
participants were included in the final analyses with 31 in the varied
group and 32, 28, 37, 25, 29, 26 participants in the constant groups
training from location 400, 500, 625, 675, 800, and 900, respectively.
All participants were compensated with course credit.

### Task and Procedure

The task of Experiment 2 was identical to that of Experiment 1, in all
but some minor adjustments to the height of the barrier, and the
relative distance between the barrier and the target. Additionally, the
intermittent testing trials featured in experiment 1 were not utilized
in experiment 2, and all training and testing trials were presented with
feedback. An abbreviated demo of the task used for Experiment 2 can be
found at (https://pcl.sitehost.iu.edu/tg/demos/igas_expt2_demo.html).

@@

### Data Processing and Statistical Packages

After confirming that condition and throwing position did not have any
significant interactions, we standardized performance within each
position, and then average across position to yield a single performance
measure per participant. This standardization did not influence our
pattern of results. As in experiment 1, we performed type III ANOVA's
due to our unbalanced design, however the pattern of results presented
below is not altered if type 1 or type III tests are used instead. The
statistical software for the primary analyses was the same as for
experiment 1. Individual learning rates in the testing phase, compared
between groups in the supplementary analyses, were fit using the TEfit
package in R (Cochrane, 2020).

### Training Phase

The different training conditions trained from positions that were not
equivalently difficult and are thus not easily amenable to comparison.
As previously stated, the primary interest of the training data is
confirmation that some learning did occur. Figure 5 depicts the training
performance of the varied group alongside that of the aggregate of the
six constant groups (5a), and each of the 6 separate constant groups
(5b). An ANOVA comparison with training stage (beginning, middle, end)
as a within-group factor and group (the varied condition vs. the 6
constant conditions collapsed together) as a between-subject factor
revealed no significant effect of group on training performance,
F(1,206)=.55,p=.49, $\eta^{2}_G$ \<.01, a significant effect of training
stage F(2,412)=77.91, p\<.001, $\eta^{2}_G$ =.05, and no significant
interaction between group and training stage, F(2,412)=.489 p=.61,
$\eta^{2}_G$ \<.01. We also tested for a difference in training
performance between the varied group and the two constant groups that
trained matching throwing positions (i.e., the constant groups training
from position 500, and position 800). The results of our ANOVA on this
limited dataset mirrors that of the full-group analysis, with no
significant effect of group F(1,86)=.48, p=.49, $\eta^{2}_G$ \<.01, a
significant effect of training stage F(2,172)=56.29, p\<.001,
$\eta^{2}_G$ =.11, and no significant interaction between group and
training stage, F(2,172)=.341 p=.71, $\eta^{2}_G$ \<.01.

@@

### Testing Phase

In Experiment 2, a single varied condition (trained from two positions,
500 and 800), was compared against six separate constant groups (trained
from a single position, 400, 500, 625, 675, 800 or 900). For the testing
phase, all participants were tested from all six positions, four of
which were novel for the varied condition, and five of which were novel
for each of the constant groups. For a general comparison, we took the
absolute deviations for each throwing position and computed standardized
scores across all participants, and then averaged across throwing
position. The six constant groups were then collapsed together allowing
us to make a simple comparison between training conditions (constant vs.
varied). A type III between-subjects ANOVA was performed, yielding a
significant effect of condition F(1,206)=4.33, p=.039, $\eta^{2}_G$
=.02. Descriptive statistics for each condition are shown in table 2.
Figure 6A visualizes the consistent advantage of the varied condition
over the constant groups across the testing positions. Figure 6b shows
performance between the varied condition and the individual constant
groups.

@@

The results of experiment 2 largely conform to the findings of
experiment 1. Participants in both varied and constant conditions
improved at the task during the training phase. We did not observe the
common finding of training under varied conditions producing worse
performance during acquisition than training under constant conditions
(Catalano & Kleiner, 1984; Wrisberg et al., 1987), which has been
suggested to relate to the subsequent benefits of varied training in
retention and generalization testing (Soderstrom & Bjork, 2015). However
our finding of no difference in training performance between constant
and varied groups has been observed in previous work (Chua et al., 2019;
Moxley, 1979; Pigott & Shapiro, 1984).

In the testing phase, our varied group significantly outperformed the
constant conditions in both a general comparison, and in an analysis
limited to novel throwing positions. The observed benefit of varied over
constant training echoes the findings of many previous visuomotor skill
learning studies that have continued to emerge since the introduction of
Schmidt's influential Schema Theory (Catalano & Kleiner, 1984; Chua et
al., 2019; Goodwin et al., 1998; McCracken & Stelmach, 1977; Moxley,
1979; Newell & Shapiro, 1976; Pigott & Shapiro, 1984; Roller et al.,
2001; Schmidt, 1975; Willey & Liu, 2018; Wrisberg et al., 1987; Wulf,
1991). We also join a much smaller set of research to observe this
pattern in a computerized task (Seow et al., 2019). One departure from
the experiment 1 findings concerns the pattern wherein the varied group
outperformed the constant group even from the training position of the
constant group, which was significant in experiment 1, but did not reach
significance in experiment 2. Although this pattern has been observed
elsewhere in the literature (Goode et al., 2008; Kerr & Booth, 1978),
the overall evidence for this effect appears to be far weaker than for
the more general benefit of varied training in conditions novel to all
training groups.

# Computational Model

Controlling for the similarity between training and testing The primary
goal of Experiment 2 was to examine whether the benefits of variability
would persist after accounting for individual differences in the
similarity between trained and tested throwing locations. To this end,
we modelled each throw as a two-dimensional point in the space of x and
y velocities applied to the projectile at the moment of release. For
each participant, we took each individual training throw, and computed
the similarity between that throw and the entire population of throws
within the solution space for each of the 6 testing positions. We
defined the solution space empirically as the set of all combinations of
x and y throw velocities that resulted in hitting the target. We then
summed each of the trial-level similarities to produce a single
similarity for each testing position score relating how the participant
threw the ball during training and the solutions that would result in
target hits from each of the six testing positions -- thus resulting in
six separate similarity scores for each participant. Figure 8a
visualizes the solution space for each location and illustrates how
different combinations of x and y velocity result in successfully
striking the target from different launching positions. As illustrated
in Figure 8b, the solution throws represent just a small fraction of the
entire space of velocity combinations used by participants throughout
the experiment.

@@

For each individual trial, the Euclidean distance (Equation 1) was
computed between the velocity components (x and y) of that trial and the
velocity components of each individual solution throw for each of the 6
positions from which participants would be tested in the final phase of
the study. The P parameter in Equation 1 is set equal to 2, reflecting a
Gaussian similarity gradient. Then, as per an instance-based model of
similarity (Logan, 2002; Nosofsky, 1992), these distances were
multiplied by a sensitivity parameter, c, and then exponentiated to
yield a similarity value. The parameter c controls the rate with which
similarity-based generalization drops off as the Euclidean distance
between two throws in x- and y-velocity space increases. If c has a
large value, then even a small difference between two throws' velocities
greatly decreases the extent of generalization from one to the other. A
small value for c produces broad generalization from one throw to
another despite relatively large differences in their velocities. The
similarity values for each training individual throw made by a given
participant were then summed to yield a final similarity score, with a
separate score computed for each of the 6 testing positions. The final
similarity score is construable as index of how accurate the throws a
participant made during the training phase would be for each of the
testing positions.

**Equation 1:** \[ Similarity\_{I,J} = \sum*{i = I}*\sum{j=J}
e^{-c^\cdot dp\_{i,j}} \]

@@

## Conclusion

Our experiments demonstrate a reliable benefit of varied training in a
simple projectile launching task. Such results were accounted for by an
instance-based model that assumes that varied training results in the
computation of a broader similarity-based generalization gradient.
Instance-based models augmented with this assumption may be a valuable
approach towards better understanding skill generalization and transfer.
